Supplementary materials

### 2021-10-18 supmat.Rmd
moderate rewording/reorganization

For many of our research questions, we wanted to know about previous work. It is common in psychology to do this via an unsystematic review -- that is, by e.g. starting out with papers one knows and then reading other papers citing those, or by doing searches on google scholar or another search engine, or by reading someone else's review. It has been well-established that overviews based on such unsystematic searches do not generalize and have other flaws because they represent biased samples of the literature [@thomas2020impact] -- for instance, famous studies are more likely to be found, even though there are no quality differences between more versus less famous studies.

So instead of relying on such an approach, we built on a systematic review that was carried out for a different paper: @cristia2020infant "combined our previous knowledge of the literature and systematic searching to yield a sample of 17 studies that we could interrogate further". Specifically, the main goal of that systematic review was to check whether "effects of infant-directed input quantities may have been reported. Input variability has not been studied, to our knowledge, but two potential proxies of input have: Monolingual status, and socio-economic status.  We set aside the literature comparing monolinguals against non-monolinguals because we reasoned that some readers may argue this contrast does not only show effects of input differences, but also interference across the languages being learned" (Cristia, 2020, personal records). Discovery combined an initial list produced by Gianmatteo Farabolini (based on readings of literature on bilingual NWR mainly), with a scholar.google.com search carried out in incognito mode by Cristia, with keywords "non-word-repetition socio-economic-status". Studies finally included reported on children who were monolingual and between 3 and 7 years of age, and which used proportion of non-words repeated as the outcome metric.

We revisited this selection of articles in order to extract data on actual performance, taking into account also the length of the non-words used in the study. From that selection, only 8 articles reported on actual performance in the text, figures, or tables, fortunately representing a range of languages: Persian by @farmani2018normalization; Israeli Arabic  [@jabere2018xperiment];  English  [@vance2005speech]; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Swedish by @kalnak2014nonword and @radeborg2006swedish. We extended that selection with 5 additional papers selected to increase the number of languages represented, with both languages sometimes described as having short words [Cantonese  @stokes2006nonword; Mandarin  @lei2011developmental]; and others described as having long words [Italian  @piazzalunga2019articulatory; Spanish @balladares2016socio]. Finally, we also included Tsimane' data  [@cristia2020infant].

@@

Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).

@@

Several observations can be drawn from this figure. To begin with, we focus on the comparison between Yélî Dnye and Tsimane'. These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects). This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech. Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate. In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process. @cristia2019child estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary. Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g. a female adult that was not temporally close to speech by others would count as child-directed). A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.

That said, @cristia2020infant also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education. This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread. If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks. Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall). These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English. What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).

Although exciting, this hypothesis is only one of many. Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study. @cristia2020infant administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead. This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise). Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.

Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies. We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores. We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources. Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall). It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g. Sotho non-words having 4-7 syllables in length.

>>>>

In order to embed our research questions in the context of previous work, we build our current study on a systematic review of NWR done for a different paper: @cristia2020infant. We preferred to build on a systematic review because, while they are common in psychology research, *unsystematic reviews*—e.g. starting out with familiar papers and then reading other papers citing those; searching via Google Scholar or another search engine; or reading someone else's review—do not generalize and represent biased samples of the literature [@thomas2020impact]. For instance, famous studies are more likely to be found, even though there are no quality differences between more versus less famous studies.

We here build on a recent and highly relevant systematic review: @cristia2020infant "combined our previous knowledge of the literature and systematic searching to yield a sample of 17 studies that we could interrogate further". Specifically, the main goal of that systematic review was to check whether "effects of infant-directed input quantities may have been reported. Input variability has not been studied, to our knowledge, but two potential proxies of input have: Monolingual status, and socio-economic status.  We set aside the literature comparing monolinguals against non-monolinguals because we reasoned that some readers may argue this contrast does not only show effects of input differences, but also interference across the languages being learned" (Cristia, 2020, personal records). Discovery combined an initial list produced by Gianmatteo Farabolini (based on readings of literature on bilingual NWR mainly), with a scholar.google.com search carried out in incognito mode by Cristia, with keywords "non-word-repetition socio-economic-status". Studies finally included reported on children who were monolingual and between 3 and 7 years of age, and which used proportion of non-words repeated as the outcome metric.

We revisited this selection of articles in order to extract data on actual performance, taking into account also the length of the non-words used in the study. From that selection, only 8 articles reported on actual performance in the text, figures, or tables, fortunately representing a range of languages: Persian by @farmani2018normalization; Israeli Arabic  [@jabere2018xperiment];  English  [@vance2005speech]; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Swedish by @kalnak2014nonword and @radeborg2006swedish. We extended that selection with 5 additional papers selected to increase the number of languages represented, with both languages sometimes described as having short words [Cantonese: @stokes2006nonword; Mandarin:  @lei2011developmental]; and others described as having long words [Italian: @piazzalunga2019articulatory; Spanish: @balladares2016socio]. Finally, we also included the Tsimane' data presented in that paper [@cristia2020infant].

@@

Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected typical word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).

@@

Several observations can be drawn from this figure. To begin with, we focus on the comparison between Yélî Dnye and Tsimane'. These two groups have been described as having (very) roughly similar lifestyles and levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects). What can we conclude from this apparent difference?

One possible lens of interpretation relates NWR differences to the minor differences in children's directed linguistic input rates. However, the differences in methods used to estimate these rates render this first interpretation unreliable. Based on current estimates Tsimane' children may encounter somewhere between similar to somewhat lower rates of directed speech comared to Yélî children. The lower boundary for Tsimane' input rate is established by @cristia2019child, who used behavioral observations of child-inclusive one-on-one conversation. The upper boundary is set by @scaff2021daylong, who used human annotation to detect speech, feeding this into an automated temporal method for assigning speech as child-directed or not (NB: this method could lead to over- but not under-estimation because any nearby speech, e.g. from a female adult, that coincided with child vocalization would count as child-directed). Yélî Dnye estimates come from @casillas2021early, who hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, effectively establishing an upper boundary of all the speech children could potentially process. There is no equivalent 'lower bound' for Yélî Dnye to compare with the Tsimane' findings. So, while it is tempting to link this difference in NWR scores to directed input rates, we will first need to actually establish input rates  using comparable methods for Yéli and Tsimane' children.

Prevalence of literacy is a second possible way to interpret these differences in NWR scores. @cristia2020infant point out the relatively low prevalence of literacy, and more generally, the variable access to formal education available in the Tsimane' population. This is very different from the present Yélî population, where most adults have accumulated some schooling, with many having basic to fluent literacy in English (note: literacy in Yélî Dnye is much less common, especially among the very young; see main text). Under this second hypothesis, there are phonetic effects of learning to read that have significant consequences for young children's encoding and decoding of sounds in the context of NWR tasks. Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance illustrated, for instance, in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall). Those two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English. What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves). However, there is at least one more plausible explanation: design and execution of the NWR task

A third, plausible, explanation for these differences in NWR is that the Tsimane' results are not generally comparable to the previous body of literature, and specifically not comparable to our present study. @cristia2020infant administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead. This implies a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise). Thus, a priority is for additional data to be gathered using this more novel testing paradigm in other populations; or more from the Tsimane' population, but using the more traditional paradigm.

Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies. We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores. We extracted the estimated number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be relatively short or long, based on information in the papers cited above and other sources. Neither of these two predictors explained the variance in average NWR scores, illustrated in Figure \@ref(fig:fig-prevlit-overall). It is possible that average word length plays a role, but that researchers tend to incorporate this into their design by including longer items when the native language allows it, with e.g. Sotho non-words having 4--7 syllables in length.
